Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-08-30 treestats_paper.tex
Add algorithms python file.
Also strip trailing spaces from tex file.
Initial draft of the branch stats algorithm.

@@

>>>>

\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\indexin}[0]{\ensuremath{\mathcal{I}}}
\newcommand{\indexout}[0]{\ensuremath{\mathcal{O}}}

@@

\clearpage
\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalised statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.}

\algstep{B1.}{Initialisation.}{
    For $0 \leq u < |\Et|$ set $\beta_u \leftarrow x_u
    \leftarrow F_u \leftarrow 0$ and $\pi_u \leftarrow -1$.
    Then, for $0 \leq j < n$ set $u \leftarrow S_j$,
    $x_u \leftarrow w_j$ and $F_u \leftarrow f(x_j)$. Finally,
    set $s \leftarrow \sigma \leftarrow j \leftarrow k \leftarrow t_l \leftarrow 0$.
}

\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$
    set $\sigma \leftarrow \sigma / L$ and terminate.
}

\algstep{B3.}{Edge removal loop.}{If $k \geq |\Et|$ or $t_l \neq
    \Et[\indexout_k]\mbox{.right}$ go to \algref{B6}}.

\algstep{B4.}{Remove edge.}{Set
    $u \leftarrow \Et[\indexout_k]\mbox{.child}$,
    $v \leftarrow \Et[\indexout_k]\mbox{.parent}$ and $k \leftarrow k + 1$.
    Then set $s \leftarrow s - \beta_u F_u$, $\pi_u \leftarrow -1$
    and $\beta_u \leftarrow 0$.
}

\algstep{B5.}{Propagate loss.}{While $v \neq -1$, set
    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v - x_u$,
    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B3}.
}

\algstep{B6.}{Edge insertion loop.}{If $j \geq |\Et|$ or $t_l \neq
    \Et[\indexin_j]\mbox{.left}$ go to \algref{B9}}.

\algstep{B7.}{Insert edge.}{Set
    $u \leftarrow \Et[\indexin_j]\mbox{.child}$,
    $v \leftarrow \Et[\indexin_j]\mbox{.parent}$ and $j \leftarrow j + 1$.
    Then set $\pi_u \leftarrow v$,
    $\beta_u \leftarrow \Nt[v]\mbox{.time} - \Nt[u]\mbox{.time}$ and
    $s \leftarrow s + \beta_u F_u$.
}

\algstep{B8.}{Propagate gain.}{While $v \neq -1$, set
    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v + x_u$,
    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B6}.
}

\algstep{B9.}{Tree loop tail.}{Set $t_r \leftarrow L$.
    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\mbox{.left})$.
    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\mbox{.right})$
    Finally, set $\sigma \leftarrow (t_r - t_l)s$,  $t_l \leftarrow t_r$
    and go to \algref{B2}.
}

\end{taocpalg}
